---
title: "Data Analytics Laboratory (CSPC-322)"
output:
  pdf_document: default
  html_document:
    df_print: paged
---

# **Lab Assignment 3**

# *Submitted by:*
<li>**Name:** Rajan Kataria</li>
<li>**Roll. No.** 18103076</li>
<li>**Branch/Sem:** CSE/6th</li>
<li>**Group:** G4</li>

## **Task:**
**Know the significance of different types of t-test**     

## **Subtasks:**         
Select a suitable dataset for following test of significance (Hypothesis testing), and interpret output parameters and their significance with respect to the selected data set and Hypothesis.        

**3.1** One sample t-test      
**3.2** Two sample t-test      
**3.3** Paired t-test           

You need to perform above given task using two different methods.    
**1.** Use inbuilt function i.e. t.test() or wilcox.test().     
**2.** Design your own functions to perform the above stated task.     

```{r}
# Importing built-in datasets library
library(datasets)
```

## **Task 3.1: One sample t-test**               

**Dataset Used:** The dataset being used for this task is the ToothGrowth dataset from built-in **datasets** package of R.

```{r}
#Understanding the structure of data
str(ToothGrowth)
```

We can see that the dataset consists of 60 observations corresponding to three attributes:        
len: Tooth Length        
supp: Supplement Type (VC - Vitamin C or OJ - Orange Juice)       
dose: Dose in milligrams/day        

To conduct the one sample t-test, we require only first attribute i.e. len.      

```{r}
# Getting subset of data for conducting one-sample t-test
one_sample_data = ToothGrowth$len
print("Population: ")
print(one_sample_data)
paste("Size of population = ",length(one_sample_data))
```


```{r}
#Getting the mean and standard deviation of the population
mean_p = mean(one_sample_data)
std_p = sd(one_sample_data)
paste("Mean of population = ",mean_p)
paste("Standard Deviation of population = ",std_p)
```


```{r}
# Getting a sample of size 20 from the population
one_sample = sample(one_sample_data, size=20, replace=F)
print("Sample: ")
print(one_sample)
```


```{r}
# Applying t-test using inbuilt function t.test()
ttest<-t.test(one_sample,mu=mean_p)
ttest
```

```{r}
names(ttest)
```


```{r}
ttest$p.value
```


```{r}
# Applying t-test using inbuilt function wilcox.test()
wilcox.test(one_sample,mu=mean_p)
```


```{r}

# Using user-defined function to conduct one-sample t-test
one_sample_t_test <- function(sampl, mu = 0, two_tailed=TRUE){
 n <- length(sampl)
 x_bar <- mean(sampl)
 s <- sd(sampl)
 t <- (x_bar - mu) / (s / sqrt(n))
 p_value <- pt(-abs(t),df=n-1)
 if(two_tailed == TRUE){
 p_value <- 2*p_value
 }

 print("Results of one-sample t-test:")
 print(paste("t = ",t," df = ",n-1," p-value = ",p_value))
 print(paste("True mean = ",mu))
 print(paste(" - NULL Hypothesis: Sample mean is equal to true mean."))
 print(paste(" - Alternate Hypothesis: Sample mean is not equal to true mean."))
 print(paste("Mean of sample = ",x_bar))
 print(paste("Standard Deviation of sample = ",s))

}
one_sample_t_test(one_sample,mu=mean_p)

```
```{r}
dim(mtcars)
```


```{r}
# One-tailed one-sample t-test
one_sample_t_test(one_sample,mu=mean_p,two_tailed=FALSE)
```


## **Task 3.2: Two sample t-test**            

**Dataset Used:** The dataset being used for this task is the ToothGrowth dataset from built-in **datasets** package of R.           

For two sample t-test, we'll compare the samples based on categories i.e. 'VC' and 'OJ'.                   

```{r}
two_sample_vc = subset(ToothGrowth,supp=="VC",len,drop=T)
two_sample_oj = subset(ToothGrowth,supp=="OJ",len,drop=T)
print("Sample 1: ")
print(two_sample_vc)
print("Sample 2: ")
print(two_sample_oj)
```


```{r}
# Performing t-test on two samples using t.test() function
t.test(two_sample_vc,two_sample_oj)

```


```{r}
# Performing t-test on two samples using wilcox.test() function
wilcox.test(two_sample_vc, two_sample_oj)
```


```{r}
# User-defined function to conduct two-sample test
two_sample_t_test <- function(sample1, sample2, two_tailed=TRUE){
 x_bar <- mean(sample1)
 y_bar <- mean(sample2)
 sx <- sd(sample1)
 sy <- sd(sample2)
 n1 <- length(sample1)
 n2 <- length(sample2)

 df <- n1+n2-2

 ss <- ((n1-1)*sx*sx + (n2-1)*sy*sy) / (n1+n2-2)
 s <- sqrt(ss)

 t <- (x_bar - y_bar) / (s * sqrt(1/n1 + 1/n2))

 p_value <- pt(-abs(t),df=df)
 if(two_tailed == TRUE){
 p_value <- 2*p_value
 }

 print("Results of two-sample t-test:")
 print(paste("t = ",t," df = ",n1+n2-2," p-value = ",p_value))

 print(paste(" - NULL Hypothesis: Difference in means of two samples is 0."))
 print(paste(" - Alternate Hypothesis: Difference in both sample means is not zero."))
 print(paste("Mean of sample 1 = ",x_bar))
 print(paste("Standard Deviation of sample 1 = ",sx))
 print(paste("Mean of sample 2 = ",y_bar))
 print(paste("Standard Deviation of sample 2 = ",sy))

}
two_sample_t_test(two_sample_vc, two_sample_oj)
```


```{r}
# Importing the dataset
life_literate <- read.csv("life_literate.csv")
head(life_literate)
```


```{r}
# Exploring the structure of data
str(life_literate)
```


## **Task 3.2: Two sample t-test**            

**Dataset Used:**For this test, we'll be using the **Life Expectancy vs Literacy Rate** by Country in 2014 dataset downloaded from https://mathcs.org/statistics/datasets/index.html


We can see above that the dataset consists of 205 observations under 3 attributes:  
Country     
Life.Expectancy      
Literacy.Rate                   

For Paired t-test, we'll consider the Life Expectancy v/s Literacy Rate of data.  


```{r}
life <- life_literate$Life.Expectancy....
literacy <- life_literate$Literacy.Rate....
print("Life Expectancy: ")
print(life)
print("Literacy Rate: ")
print(literacy)
```


```{r}
# Paired t-test using t.test() function
t.test(life, literacy, paired = TRUE)
```


```{r}
# Paired t-test using wilcox.test() function
wilcox.test(life, literacy, paired = TRUE)
```


```{r}
# User-defined function to conduct paired t-test
paired_t_test <- function(X, Y, two_tailed=TRUE){
 D <- X-Y
 d_bar = mean(D)
 std_d = sd(D)
 n = length(D)
 df = n-1

 t <- d_bar / (std_d / sqrt(n))

 p_value <- pt(-abs(t),df=df)
 if(two_tailed == TRUE){
 p_value <- 2*p_value
 }

 print("Results of paired t-test:")
 print(paste("t = ",t," df = ",df," p-value = ",p_value))

 print(paste(" - NULL Hypothesis: Difference in means of two samples is 0."))
 print(paste(" - Alternate Hypothesis: Difference in both sample means is not zero."))
 print(paste("Mean of differences = ",d_bar))
 print(paste("Standard Deviation of differences = ",std_d))
}
paired_t_test(life, literacy)
```


#### Assign 4

```{r}
crop.df<-read.csv('crop.data.csv')
```

```{r}
dim(crop.df)
```


```{r}
str(crop.df)
```


```{r}
head(crop.df)
```
```{r}
table(crop.df$density)
```


```{r}
library(dplyr)
crop.df %>% summarize_all(n_distinct)
```

